---
title: "US Census"
output:
  html_document:
    df_print: paged
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(tigris)
library(sf)
```

The `tigris` package defines functions that will download spatial data. It uses the `sp` classes, so we convert to `sf` using `st_as_sf` and then transform to Texas Centric Albers Equal Area projection. We do not use the full resolution data here, which would be obtained if we did not specify the `cb` and `resolution` switches.

```{r results='hide'}
texas.counties <- counties(48, cb = TRUE, resolution = '5m') %>%
  st_as_sf() %>%
  st_transform(crs = 3083)
```

Now we can plot the result.

```{r}
ggplot(texas.counties) +
  geom_sf(aes(fill = rank(as.integer(AWATER)))) +
  scale_fill_gradient(low = "tan", high = "white") +
  guides(fill = FALSE) +
  theme_bw()
```

Here's an example from the author of the `tigris` package. First we retrieve some party affiliation data (certainly dated) from a website. Then we join that data to the spatial data using `dplyr`'s `inner_join` function.

```{r results='hide'}
party <- read_csv("http://personal.tcu.edu/kylewalker/data/txlege.csv", col_types = cols())
districts <- state_legislative_districts("TX", house = "lower", cb = TRUE) %>%
  st_as_sf() %>%
  st_transform(crs = 3083) %>%
  inner_join(y = party, by = c("NAME" = "District")) %>%
  mutate(Party = as.factor(recode(Party, R = "Republican", D = "Democrat")))
```

Now we have a spatial coverage where each polygon (lower house districts in Texas) has a party affiliation. Let's make a plot.

```{r}
ggplot(districts) +
  geom_sf(aes(color = Party, fill = Party), alpha = 0.5) +
  scale_fill_manual(values = c("blue", "red")) +
  scale_color_manual(values = c("darkblue", "darkred")) +
  theme_bw()
```
