---
title: "Mapping Texas Parks"
output:
  html_document:
    df_print: paged
---

This is an example of plotting data from TXPWD. The Gould Eco Regions data can be found [here](https://tpwd.texas.gov/gis/resources/gould-eco-regions.zip).

Load ```sf```.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
```

Read in the shapefile.

```{r}
ger <- st_read("GouldEcoRegions.shp") %>% st_simplify(dTolerance = 1e3)
```

Print summary of the columns.

```{r}
summary(ger)
```

Look the Coordinate Reference System. Its the [Texas State Mapping System](http://epsg.io/3081-1734).

```{r}
st_crs(ger)
```

We can plot using the ```sf``` plot function.

```{r}
if (compareVersion(as.character(packageVersion("ggplot2")), "2.2.1") < 1)
  devtools::install_github("hadley/ggplot2")
library(ggplot2)
```

```{r}
ggplot() +
  geom_sf(data = ger, aes(fill = Shape_Area)) -> er.plot
er.plot
```

```{r}
er.plot + geom_sf(data = st_centroid(ger), color = "red")
```

Read in park boundaries

```{r}
parks <- st_read("TPWDStateParksBoundary.shp") %>% st_simplify(dTolerance = 1e3)
```

With buffering and alpha, we can create a halo around parks to make small polygons more visible.

```{r}
er.plot +
  geom_sf(data = st_buffer(parks, 2e4), color = NA, fill = "darkred", alpha = 0.25) +
  geom_sf(data = parks, color = "red", fill = NA)
```