```{r include=F}
#opts_chunk$set(warning=F, message=F)

solutions=F

```

Load a few plot helpers.

```{r}
color.ramp <-function(palette="Spectral", reverse=TRUE, sample=FALSE)
{
  pacman::p_load(RColorBrewer)
  n <- brewer.pal.info[palette,]$maxcolors
  pal <- brewer.pal(n, palette)
  if (reverse) pal <- rev(pal)
  if (sample) pal <- sample(pal)
  return(colorRampPalette(pal))
}

set.spplot.colors <- function(ncol = 100, palette = "Spectral", reverse = TRUE, sample = FALSE)
{
  pacman::p_load(lattice)
  pal <- color.ramp(palette, reverse, sample)(ncol)
  trellis.par.set(regions = list(col = pal))
}
```

# Geospatial Data Analysis in `R`
## Raster Data Sources

```{r}
pacman::p_load(raster)
pacman::p_load(rasterVis)
```

### Raster Visualization
```{r}
r <- raster("ncrast/elev_lid792_1m.tif")
```

#### Visualizing Values
```{r}
contourplot(r)
levelplot(r, col.regions=terrain.colors)
```

#### Visualizing How Values are Changing
```{r}
vectorplot(r, col.regions=terrain.colors)
streamplot(r)
```

#### Interactive 3D Visualization
Mouse wheel to zoom, click and drag to rotate. This code will not be run when generating the html. Try it in R.
```{r eval=F}
pacman::p_load("rasterVis")
r <- raster("ncrast/elev_lid792_1m.tif")
plot3D(r)
```

### Working With Raster Values

#### Convert to Matrix
```{r}
m <- as.matrix(r)
m[1:3, 1:3]         # first 3 rows and columns
```

#### Aggregate
```{r}
r.mean <- aggregate(r, fact=10, fun=mean)
r.max <- aggregate(r, fact=10, fun=max)
r.min <- aggregate(r, fact=10, fun=min)
```

Dimensions changed:
```{r results='hold'}
dim(r)
dim(r.mean)
```

Spatial extent unchanged:
```{r results='hold'}
extent(r)
extent(r.mean)
```

#### Disaggregate
Either repeats values or linearly interpolates them.
```{r}
r.mean <- disaggregate(r.mean, fact=10)
r.mean.interp <- disaggregate(r.max, fact=10, method='bilinear')
```
Raster layers must have identical spatial extent and resolution to be stacked.
```{r}
levelplot(stack(r, r.mean, r.mean.interp), col.regions=terrain.colors)
```

### Raster to Vector Conversion
#### Raster to XYZ Points
The Z coordinate is the raster value.
```{r}
pts <- rasterToPoints(r)
head(pts)
```

#### Raster to `SpatialPoints`
```{r}
sp.pts <- rasterToPoints(r, spatial=T)
summary(sp.pts)
```

#### Raster to `SpatialPolygons`
The `raster` package provides `rasterToPolygons` but performance is unacceptable for large rasters. Use the command line utility `gdal_polygonize.py` instead. In the "tools" menu above in RStudio, you will see a "shell" option. You can use that to enter these commands. (Your directory probably already has these files, so you may not need to do this.)
```
cd Exercises
gdal_polygonize.py ncrast/zipcodes.tif -f "ESRI Shapefile" extra zipcodes zipnum
```
Now read in the shapefile and convert the zipnum attribute to a factor:
```{r}
pacman::p_load(rgdal)
zipcodes <- readOGR(dsn="extra", layer="zipcodes", verbose=F)
zipcodes$zipnum <- factor(zipcodes$zipnum)
spplot(zipcodes, col.regions=color.ramp("Set3")(nlevels(zipcodes$zipnum)))
```

#### Activity: Reproject a Raster
Use the `projectRaster` function to reproject the *elev_state_500m.tif* elevation map to longitude/latitude coordinates. Call `projectRaster` with a raster object and a named crs string `crs=+proj=longlat`. Then usen `levelplot` to visualize the result.

Functions: `raster`, `projectRaster`, `levelplot`

```{r echo=solutions, fig.width=10}
r <- raster("ncrast/elev_state_500m.tif")
s <- projectRaster(r, crs="+proj=longlat")
levelplot(s, col.regions=terrain.colors)
```

#### Activity: Calculate NDVI
Use the near infrared (*lsat7_2002_40.tif*) and visible (*lsat7_2002_30.tif*) LSAT imagery to calculat NDVI. NDVI is defined as (NIR - VIS)/(NIR + VIS). Use `levelplot` to visualize the result.

Functions: `raster`, `levelplot`, `rev`, `terrain.colors`

```{r echo=solutions, fig.width=10}
nir <- raster("ncrast/lsat7_2002_40.tif")
vis <- raster("ncrast/lsat7_2002_30.tif")
ndvi <- (nir - vis) / (nir + vis)
levelplot(ndvi, col.regions=rev(terrain.colors(20)))
```